---
title: "Predicting Political Violence in the United States"
format: html
editor: visual
---

# Predicting Political Violence in the United States Code

```{r}
library(tidyverse)
library(regclass)
library(lmtest)
library(ResourceSelection)
library(glmnet)
library(pROC)
```

# Data Exploration and Preparation

## Data Preparation

In order to make sure that I was able to train my models on the best subset of data possible, I had to read through the PIRUS Codebook to develop an understanding of what each variable in the dataset meant, and consider what that variable might mean for my models. For starters, I filtered out all observations that held "NA" values. Of the 124 variables in the original, only 10 are numeric, and each variable that held "NA" values was a categorical variable. Performing imputation on these variables would have been difficult and could have negatively influenced my model, so for the time being I figured it was best to just remove them from the training data. I also removed variables related to the violent plots of the extremists, as these values would have been directly informative of my response variable. Lastly, I removed variables like SubjectID and the social network nodes, as these are irrelevant to the goal of predicting whether or not an extremist will participate in or commit an act of political violence.

```{r}
pirus <- read.csv("pirus_encoded.csv")
glimpse(pirus)
# When reading in my encoded file, it reads all the factors in as character vectors.
pirus_fct <- pirus %>% 
  mutate(across(-c(Subject_ID, Loc_Plot_State1:Year_Exposure, Loc_Habitation_State1:Loc_Habitation_City2, Age, Children, Citizenship, Nativity, Immigrant_Source, Component_ID, Component_Density, Degree_Centrality:Closeness_Centrality), as.factor))
  
glimpse(pirus_fct)
# Filtering and cleaning this dataset is going to be an absolute headache. In order to make this more readable, I'm going to remove "groups" of columns at a time.
political_extremism <- pirus_fct %>%
  filter(Radicalization_Islamist != "Yes") %>%
  filter(Ideological_Sub_Category1 != "Christian Identity" & Ideological_Sub_Category1 != "Islamist" & Ideological_Sub_Category1 != "Jewish Defense Lg.") %>% # Because I'm just focusing on politically-motivated extremism and not religiously-motivated extremism, I'm removing people with religious ideological sub categories.
  select(-c(Subject_ID, Loc_Plot_State1:Loc_Plot_City2, Plot_Target1:Extent_Plot, Primary_Event_Mass_Casualty_Incident:Foreign_Fighter, Component_ID, Component_Density, Degree_Centrality:Closeness_Centrality)) %>% 
  mutate(Violent = case_when(
# Have to make Violent a binomial variable for a logistic regression
    Violent == "Yes" ~ 1,
    Violent == "No" ~ 0
  ))# This removes all of the variables related to plots and the social network nodes of an extremist
```

There are some variables that are largely filled with NA values, so I'm going to exclude these from my maximum model. Filtering for instances where these values aren't NA will cut down my dataset significantly.

```{r}
political_extremism <- political_extremism %>% 
  select(-c(Criminal_Severity:Current_Status, Recruiter2, Recruiter3, Social_Media_Platform2:Social_Media_Platform5, Social_Media_Activities2:Social_Media_Activities7, Loc_Habitation_State1:Itinerant, Event_Influence2:Event_Influence4, Abuse_Type2, Abuse_Type3, Previous_Criminal_Activity_Type2, Previous_Criminal_Activity_Type3, Nativity, Citizenship, Time_US_Months, Radicalization_Islamist, Ideological_Sub_Category2, Ideological_Sub_Category3)) %>%
## Making one Radicalization Variable
  mutate(Radicalization = case_when(
    Radicalization_Far_Right == "Yes" ~ "Far Right",
    Radicalization_Far_Left == "Yes" ~ "Far Left",
    Radicalization_Single_Issue == "Yes" ~ "Single Issue"
  ),
  Radicalization = as.factor(Radicalization),
  Violent = as.factor(Violent)) %>% 
  select(-Radicalization_Far_Right, -Radicalization_Far_Left, -Radicalization_Single_Issue) %>%
  # filtering out any person who's age is unknown
  filter(Age != -99) %>%
  drop_na()

glimpse(political_extremism)
```

## Data Exploration

```{r}
# Because most of the variables in this dataset are categorical, there are only a couple variables of interest I can take summary statistics of.
summary(political_extremism$Year_Exposure)
summary(political_extremism$Age)

# Looking at the earliest and most recent observations
political_extremism %>% 
  select(Year_Exposure, Violent, Age, Ideological_Sub_Category1, Radicalization) %>% 
  filter(Year_Exposure == min(Year_Exposure) | Year_Exposure == max(Year_Exposure)) %>% 
  group_by(Year_Exposure) %>% 
  slice_head(n = 1)
```

```{r}
# This next plot can give me a good idea of how many observations I am working with for my response variable, Violent
ggplot(political_extremism, aes(x = Violent)) +
  geom_bar(stat = "count", fill = "firebrick") +
  labs(title = "Distribution of Exposed Non-Violent and Violent Extremists in PIRUS",
       x = "Violent (0 = No, 1 = Yes)",
       y = "Number of Extremists") +
  set_theme(new = theme_minimal())
```

```{r}
# Making a summary table of total extremists by age group
political_extremism %>% 
  mutate(Age_Group = case_when(
    Age < 10 ~ "0-10",
    Age < 20 ~ "10-20",
    Age < 30 ~ "20-30",
    Age < 40 ~ "30-40",
    Age < 50 ~ "40-50",
    Age < 60 ~ "50-60",
    Age < 70 ~ "60-70",
    Age < 80 ~ "70-80",
    Age < 90 ~ "80-90",
    Age < 100 ~ "90-100"
  )) %>% 
  group_by(Age_Group) %>% 
  summarise(Total = n())

# Plotting the above table
political_extremism %>% 
  mutate(Age_Group = case_when(
    Age < 10 ~ "0-10",
    Age < 20 ~ "10-20",
    Age < 30 ~ "20-30",
    Age < 40 ~ "30-40",
    Age < 50 ~ "40-50",
    Age < 60 ~ "50-60",
    Age < 70 ~ "60-70",
    Age < 80 ~ "70-80",
    Age < 90 ~ "80-90",
    Age < 100 ~ "90-100"
  )) %>% 
  ggplot(aes(x = Age_Group, fill = Violent)) +
  geom_bar(position = "dodge") + 
  labs(title = "Distribution of Exposed Non-Violent and Violent Extremists in PIRUS by Age Group",
       x = "Age Group",
       y = "Number of Extremists",
       fill = "Violent\n(0 = No, 1 = Yes)") +
  scale_fill_manual(values = c("forestgreen", "firebrick")) +
  set_theme(new = theme_minimal())
```

```{r}
# plotting the number of exposed violent and non-violent extremists over time
political_extremism %>% 
  group_by(Violent, Year_Exposure) %>% 
  summarise(Total = n()) %>% 
  ggplot(aes(x = Year_Exposure, y = Total, color = Violent, group = Violent)) +
  geom_line() +
  labs(title = "Number of Exposed Violent and Non-Violent Extremists Over Time",
       x = "Year",
       y = "Number of Extremists",
       color = "Violent\n(0 = No, 1 = Yes)") +
  scale_color_manual(values = c("forestgreen", "firebrick")) + 
  set_theme(new = theme_minimal())
```

# Model Building

## Building an Unpenalized Logistic Regression Model

Splitting my data with an 80-20 split.

```{r}
set.seed(2212003)
Violent <- political_extremism$Violent
political_extremism_exp <- political_extremism %>% 
  select(-Violent)

pirus_matrix <- model.matrix( ~ -1 + ., political_extremism_exp)
pirus2 <- as.data.frame(pirus_matrix)
pirus2 <- cbind(pirus2, Violent)

train_index <- sample(1:nrow(pirus2), size = 0.8 * nrow(pirus2))
train_data <- pirus2[train_index,]
test_data <- pirus2[-train_index,]
# Dropping factor levels not in the data to avoid multicollinearity
train_data <- droplevels(train_data)
test_data <- droplevels(test_data)

# Making train and test data
x_train <- model.matrix(Violent ~ ., train_data)[, -1]
y_train <- train_data$Violent

x_test <- model.matrix(Violent ~ ., test_data)[, -1]
y_test <- test_data$Violent
```

```{r}
# Up first: logistic regression with all variables
mod_max <- glm(Violent ~ ., train_data, family = "binomial")
summary(mod_max)
```

I'll be looking at the **Area Under the Curve (AUC)** and **test accuracy** of my models to select my best model, and I will be **comparing the train accuracy to the test accuracy** of each model to determine if the models fit my data well.

```{r}
# Testing on the train data first
probs_max_train <- predict(mod_max, newdata = train_data, type = "response")
pred_max_train <- ifelse(probs_max_train > 0.5, 1, 0)
max_train_acc <- mean(pred_max_train == y_train)
max_train_acc

# Now on the test data
probs_max_test <- predict(mod_max, newdata = test_data, type = "response")

plot(roc(y_test, probs_max_test), main = "Unpenalized Logistic Regression")
auc(roc(y_test, probs_max_test))

pred_max_test <- ifelse(probs_max_test > 0.5, 1, 0)
max_test_acc <- mean(pred_max_test == y_test)
max_test_acc

mod_max_coefs <- as.data.frame(summary(mod_max)$coefficients)
mod_max_coefs
```

AUC = 0.8586 \~ 85.9%, Train Accuracy = 0.8949 \~ 89.5%, Test Accuracy = 0.7944 \~ 79.4%

## Building LASSO Regression Models with K-Fold Cross Validation

```{r}
# Training the model
cv_lasso <- cv.glmnet(x = x_train, y = y_train, nfolds = 10, alpha = 1, family = "binomial")
plot(cv_lasso)
lasso_min <- cv_lasso$lambda.min
lasso_1se <- cv_lasso$lambda.1se

lasso_min_mod <- glmnet(x_train, y_train, lambda = lasso_min, alpha = 1, family = "binomial")
lasso_1se_mod <- glmnet(x_train, y_train, lambda = lasso_1se, alpha = 1, family = "binomial")

coef(lasso_min_mod)
lasso_1se_coefs <- coef(lasso_1se_mod)
```

```{r}
# Testing my models, starting with min
probs_lmin_train <- predict(lasso_min_mod, newx = x_train, s = "lambda.min", type = "response")
pred_lmin_train <- ifelse(probs_lmin_train > 0.5, 1, 0)
lmin_train_acc <- mean(pred_lmin_train == y_train)
lmin_train_acc

probs_lmin_test <- predict(lasso_min_mod, newx = x_test, s = "lambda.min", type = "response")

plot(roc(y_test, probs_lmin_test), main = "LASSO Regression with lambda.min")
auc(roc(y_test, probs_lmin_test))

pred_lmin_test <- ifelse(probs_lmin_test > 0.5, 1, 0)
lmin_test_acc <- mean(pred_lmin_test == y_test)
lmin_test_acc
```

The AUC is 0.9242 \~ 92.4%, the train accuracy is 0.8625 \~ 86.3%, and the test accuracy is 0.8481 \~ 84.8%.

```{r}
# Now for se1
probs_l1se_train <- predict(lasso_1se_mod, newx = x_train, s = "lambda.1se", type = "response")
pred_l1se_train <- ifelse(probs_l1se_train > 0.5, 1, 0)
l1se_train_acc <- mean(pred_l1se_train == y_train)
l1se_train_acc

probs_l1se_test <- predict(lasso_1se_mod, newx = x_test, s = "lambda.1se", type = "response")

plot(roc(y_test, probs_l1se_test), main = "LASSO Regression with lambda.1se")
auc(roc(y_test, probs_l1se_test))

pred_l1se_test <- ifelse(probs_l1se_test > 0.5, 1, 0)
l1se_test_acc <- mean(pred_l1se_test == y_test)
l1se_test_acc
```

The AUC of this model is 0.9305 \~ 93.1%, the train accuracy is 0.8421 \~ 84.2%, and the test accuracy is 0.8574 \~ 85.7%.

## Building Ridge Regression Models with K-Fold Cross Validation

```{r}
cv_ridge <- cv.glmnet(x = x_train, y = y_train, nfolds = 10, alpha = 0, family = "binomial")
plot(cv_ridge)

ridge_min <- cv_ridge$lambda.min
ridge_1se <- cv_ridge$lambda.1se

ridge_min_mod <- glmnet(x_train, y_train, lambda = ridge_min, alpha = 0, family = "binomial")
ridge_1se_mod <- glmnet(x_train, y_train, lambda = ridge_1se, alpha = 0, family = "binomial")

ridge_min_coefs <- coef(ridge_min_mod)
ridge_1se_coefs <- coef(ridge_1se_mod)
```

```{r}
# Testing on train data
probs_rmin_train <- predict(ridge_min_mod, newx = x_train, s = "lambda.min", type = "response")
pred_rmin_train <- ifelse(probs_rmin_train > 0.5, 1, 0)
rmin_train_acc <- mean(pred_rmin_train == y_train)
rmin_train_acc

# Testing on test data with ROC/AUC
probs_rmin_test <- predict(ridge_min_mod, newx = x_test, s = "lambda.min", type = "response")

plot(roc(y_test, probs_rmin_test), main = "Ridge Regression with lambda.min")
auc(roc(y_test, probs_rmin_test))

pred_rmin_test <- ifelse(probs_rmin_test > 0.5, 1, 0)
rmin_test_acc <- mean(pred_rmin_test == y_test)
rmin_test_acc
```

The AUC of this model is 0.9008 \~ 90.1%, the train accuracy is 0.8773 \~ 87.7%, and the test accuracy is 0.8166 \~ 81.7%

```{r}
# Train first
probs_r1se_train <- predict(ridge_1se_mod, newx = x_train, s = "lambda.1se", type = "response")
pred_r1se_train <- ifelse(probs_r1se_train > 0.5, 1, 0)
r1se_train_acc <- mean(pred_r1se_train == y_train)
r1se_train_acc

# Test second
probs_r1se_test <- predict(ridge_1se_mod, newx = x_test, s = "lambda.1se", type = "response")
# AUC/ROC
plot(roc(y_test, probs_r1se_test), main = "Ridge Regression with lambda.1se")
auc(roc(y_test, probs_r1se_test))

pred_r1se_test <- ifelse(probs_r1se_test > 0.5, 1, 0)
r1se_test_acc <- mean(pred_r1se_test == y_test)
r1se_test_acc
```

The AUC for this model is 0.9019 \~ 90.2%, the train accuracy is 0.8685 \~ 86.9%, and the test accuracy is 0.8222 \~ 82.2%.

Now, for the last models, I will build elastic nets.

## Building Elastic Net Regression Models with K-Fold Cross Validation

```{r}
cv_enet <- cv.glmnet(x = x_train, y = y_train, nfolds = 10, alpha = 0.5, family = "binomial")
plot(cv_enet)

enet_min <- cv_enet$lambda.min
enet_1se <- cv_enet$lambda.1se

enet_min_mod <- glmnet(x_train, y_train, lambda = enet_min, alpha = 0.5, family = "binomial")
enet_1se_mod <- glmnet(x_train, y_train, lambda = enet_1se, alpha = 0.5, family = "binomial")

enet_min_coefs <- coef(enet_min_mod)
enet_1se_coefs <- coef(enet_1se_mod)
```

Generating my predictions for the elastic net models, starting with the lambda min model

```{r}
# Train
probs_emin_train <- predict(enet_min_mod, newx = x_train, s = "lambda.min", type = "response")
pred_emin_train <- ifelse(probs_emin_train > 0.5, 1, 0)
emin_train_acc <- mean(pred_emin_train == y_train)
emin_train_acc

# Test
probs_emin_test <- predict(enet_min_mod, newx = x_test, s = "lambda.min", type = "response")

plot(roc(y_test, probs_emin_test), main = "Elastic Net with lambda.min")
auc(roc(y_test, probs_emin_test))

pred_emin_test <- ifelse(probs_emin_test > 0.5, 1, 0)
emin_test_acc <- mean(pred_emin_test == y_test)
emin_test_acc
```

The AUC of this model is 0.9239 \~ 92.4%, the train accuracy is 0.868 \~ 86.7%, and the test accuracy is 0.8407 \~ 84.1%

```{r}
probs_e1se_train <- predict(enet_1se_mod, newx = x_train, s = "lambda.1se", type = "response")
pred_e1se_train <- ifelse(probs_e1se_train > 0.5, 1, 0)
e1se_train_acc <- mean(pred_e1se_train == y_train)
e1se_train_acc

probs_e1se_test <- predict(enet_1se_mod, newx = x_test, s = "lambda.1se", type = "response")

plot(roc(y_test, probs_e1se_test), main = "Elastic Net with lambda.1se")
auc(roc(y_test, probs_e1se_test))

pred_e1se_test <- ifelse(probs_e1se_test > 0.5, 1, 0)
e1se_test_acc <- mean(pred_e1se_test == y_test)
e1se_test_acc
```

The AUC of this model is 0.927 \~ 92.7%, the train accuracy is 0.8504 \~ 85%, and the test accuracy is 0.8388 \~ 83.9%.

# Interpretation of Results

## Model Diagnostic Results Table

| Model | \# of Coefficients | AUC | Train Accuracy | Test Accuracy | Accuracy Difference |
|-------------|------------|------------|------------|------------|---------------|
| Unpenalized Logistic Regression | 342 (24 undef.) | *0.8586 (85.9%)* | **0.8949 (89.5%)** | *0.7944 (79.4%)* | *-0.1005 (10.1%)* |
| LASSO Logistic Regression ($\lambda = \hat{\lambda}$) | 85 | 0.9242 (92.4%) | 0.8625 (86.3%) | 0.8481 (84.8%) | -0.0144 (1.4%) |
| LASSO Logistic Regression ($\lambda = \bar{\lambda}$) | 38 | **0.9305 (93.1%)** | 0.8421 (84.2%) | **0.8574 (85.7%)** | **+0.0153 (+1.5%)** |
| Ridge Logistic Regression ($\lambda = \hat{\lambda}$) | 350 | 0.9008 (90.1%) | 0.8773 (87.7%) | 0.8166 (81.7%) | -0.0607 (6.1%) |
| Ridge Logistic Regression ($\lambda = \bar{\lambda}$) | 350 | 0.9019 (90.2%) | 0.8685 (86.9%) | 0.8222 (82.2%) | -0.0463 (4.6%) |
| Elastic Net Logistic Regression ($\lambda = \hat{\lambda}$) | 111 | 0.9239 (92.3%) | 0.8680 (86.8%) | 0.8407 (84.1%) | -0.0273 (2.7%) |
| Elastic Net Logistic Regression ($\lambda = \bar{\lambda}$) | 52 | 0.927 (92.7%) | 0.8504 (85%) | 0.8388 (83.9%) | -0.0116 (1.2%) |

: **Highest is bold**, *lowest is italicized*

## Interpretations from my Best Model

```{r}
mod_best_df <- as.data.frame(as.matrix(lasso_1se_coefs)) %>% 
  filter(s0 != 0)
mod_best_df
```

There are 38 coefficients in the LASSO Logistic Regression model with $\lambda = \bar{\lambda}$

**Intercept:** $\text{log}(\hat{Violent}) = -6.653$, exp(-6.653) = 0.00129

-   The odds that a political extremist will participate in, or commit an act of political violence are 0.00129, holding all numeric explanatory variables at 0 and all categorical explanatory variables at their baseline category.

$\beta_{22} RadicalBehaviors ViolentPlot = 3.281$, exp(3.281) = 26.6:

-   The odds that an extremist will participate in or commit an act of political violence are 26.6 times higher for an extremist who creates or contributes to a violent compared to an extremist who simply associates with radicals, holding all other variables constant.

This variable deserves some discussion. At first when I saw this variable and interpreted it, I was worried that it should have been removed originally when I removed the variables related to plots. I looked at the original PIRUS dataset, and saw that all of the violent plot variables I removed were perfectly indicative of a political extremist committing an act of violence. Radical Behaviors Violent Plot, however, is not. Roughly one third of the extremists who did commit an act of political violence observed in this dataset did not plot or contribute to a violent plot, and there are a small amount (roughly 3%) of the non-violent extremists who did plot or contribute to a violent plot, but never actually committed or participated in an act of political violence. In further research, I could toy around with removing it or filtering out the observations where Radical Behaviors is equal to Violent Plot, but for now, I think I should keep it in.

$\beta_{26} SexMale = 0.1196$, exp(0.1196) = 1.126:

-   The odds that an extremist will participate in or commit an act of political violence are 1.126 times higher for a male extremist than a female extremist, holding all other variables constant.

Unfortunately, due to the way my models were trained, I cannot run the glmnet predict function on new data that doesn't hold values for all 365 explanatory variables in my train data. However, hopefully as new data is added to the PIRUS dataset, I will be able to run my models again and see similar results as to what I got during this project. All of my models, with the exception of the unpenalized logistic regression model, performed exceptionally well on the test dataset, so I have reason to believe that as new data is added, I will see similar results. Being able to have a handful of models with AUC probabilities of greater than 90% demonstrates the predictive power that can be derived from implementing regularization techniques.
