---
title: "PIRUS"
format: html
editor: visual
---

# Political Extremism Regression Project

## Introduction

For this project, I wanted to build a logistic regression model that could predict whether or not a Far-Left, Far-Right, or Single–Issue radicalized individual in the United States would plot or commit a physically violent act. The dataset that I used to train this model is the Profiles of Individuals Radicalized in the United States (PIRUS) dataset from the University of Maryland's National Consortium for the Study of Terrorism and Responses to Terrorism (START). This dataset contains information about extremist plots, activities, radicalization sequences and details, demographics, and social network nodes. The original dataset has 3,526 observations which span from 1940-2022. Each observation is deidentified, but several notable domestic terrorist attacks, such as the Oklahoma City Bombing in 1994, the Charleston Church shooting in 2015, the El Paso Walmart shooting in 2019, and the Buffalo Grocery Store shooting in 2022. What is notably excluded from PIRUS is the January 6th Insurrection at the United States Capitol Building in 2021, which was excluded because of the high number of perpetrators (\~1000), which would drastically influence any models or research on this dataset. I chose to do this project because of it's relevance to the current political climate in the United States. Extremism influenced by political leaning has seen a rise in the 21st Century, and violent acts motivated by these extreme beliefs have also seen a rise. The publicity of this extremism has also increased, with media outlets and politicians spinning extremist activities every which way in order to fit their narrative. My hope is that the predictive models I have built can cut through these narratives and deliver real, unbiased insights into what pushes political extremists to the violent edge.

```{r}
library(tidyverse)
library(regclass)
library(lmtest)
library(ResourceSelection)
library(glmnet)
library(pROC)
```

# Data Exploration and Preparation

## Data Preparation

In order to make sure that I was able to train my models on the best subset of data possible, I had to read through the PIRUS Codebook to develop an understanding of what each variable in the dataset meant, and consider what that variable might mean for my models. For starters, I filtered out all observations that held "NA" values. Of the 124 variables in the original, only 10 are numeric, and each variable that held "NA" values was a categorical variable. Performing imputation on these variables would have been difficult and could have negatively influenced my model, so for the time being I figured it was best to just remove them from the training data. I also removed variables related to the violent plots of the extremists, as these values would have been directly informative of my response variable. Lastly, I removed variables like SubjectID and the social network nodes, as these are irrelevant to the goal of predicting whether or not an extremist will plot or commit a physically violent act. For

```{r}
pirus <- read.csv("pirus_encoded.csv")
glimpse(pirus)
# When reading in my encoded file, it reads all the factors in as character vectors.
pirus_fct <- pirus %>% 
  mutate(across(-c(Subject_ID, Loc_Plot_State1:Year_Exposure, Loc_Habitation_State1:Loc_Habitation_City2, Age, Children, Citizenship, Nativity, Immigrant_Source, Component_ID, Component_Density, Degree_Centrality:Closeness_Centrality), as.factor))
  
glimpse(pirus_fct)
# Filtering and cleaning this dataset is going to be an absolute headache. In order to make this more readable, I'm going to remove "groups" of columns at a time.
political_extremism <- pirus_fct %>%
  filter(Radicalization_Islamist != "Yes") %>%
  filter(Ideological_Sub_Category1 != "Christian Identity" & Ideological_Sub_Category1 != "Islamist" & Ideological_Sub_Category1 != "Jewish Defense Lg.") %>% # Because I'm just focusing on politically-motivated extremism and not religiously-motivated extremism, I'm removing people with religious ideological sub categories.
  select(-c(Subject_ID, Loc_Plot_State1:Loc_Plot_City2, Plot_Target1:Extent_Plot, Primary_Event_Mass_Casualty_Incident:Foreign_Fighter, Component_ID, Component_Density, Degree_Centrality:Closeness_Centrality)) %>% 
  mutate(Violent = case_when(
# Have to make Violent a binomial variable for a logistic regression
    Violent == "Yes" ~ 1,
    Violent == "No" ~ 0
  ))# This removes all of the variables related to plots and the social network nodes of an extremist
```

There are some variables that are largely filled with NA values, so I'm going to exclude these from my maximum model. Filtering for instances where these values aren't NA will cut down my dataset significantly.

```{r}
political_extremism <- political_extremism %>% 
  select(-c(Criminal_Severity:Current_Status, Recruiter2, Recruiter3, Social_Media_Platform2:Social_Media_Platform5, Social_Media_Activities2:Social_Media_Activities7, Loc_Habitation_State1:Itinerant, Event_Influence2:Event_Influence4, Abuse_Type2, Abuse_Type3, Previous_Criminal_Activity_Type2, Previous_Criminal_Activity_Type3, Nativity, Citizenship, Time_US_Months, Radicalization_Islamist, Ideological_Sub_Category2, Ideological_Sub_Category3)) %>%
## Making one Radicalization Variable
  mutate(Radicalization = case_when(
    Radicalization_Far_Right == "Yes" ~ "Far Right",
    Radicalization_Far_Left == "Yes" ~ "Far Left",
    Radicalization_Single_Issue == "Yes" ~ "Single Issue"
  ),
  Radicalization = as.factor(Radicalization),
  Violent = as.factor(Violent)) %>% 
  select(-Radicalization_Far_Right, -Radicalization_Far_Left, -Radicalization_Single_Issue) %>%
  # filtering out any person who's age is unknown
  filter(Age != -99) %>% 
  drop_na()

glimpse(political_extremism)
```

## Data Exploration

```{r}
# Because most of the variables in this dataset are categorical, there are only a couple variables of interest I can take summary statistics of.
summary(political_extremism$Year_Exposure)
summary(political_extremism$Age)

# Looking at the earliest and most recent observations
political_extremism %>% 
  select(Year_Exposure, Violent, Age, Ideological_Sub_Category1, Radicalization) %>% 
  filter(Year_Exposure == min(Year_Exposure) | Year_Exposure == max(Year_Exposure)) %>% 
  group_by(Year_Exposure) %>% 
  slice_head(n = 1)
```

```{r}
# This next plot can give me a good idea of how many observations I am working with for my response variable, Violent
ggplot(political_extremism, aes(x = Violent)) +
  geom_bar(stat = "count", fill = "firebrick") +
  labs(title = "Distribution of Exposed Non-Violent and Violent Extremists in PIRUS",
       x = "Violent (0 = No, 1 = Yes)",
       y = "Number of Extremists") +
  set_theme(new = theme_minimal())
```

```{r}
# Making a summary table of total extremists by age group
political_extremism %>% 
  mutate(Age_Group = case_when(
    Age < 10 ~ "0-10",
    Age < 20 ~ "10-20",
    Age < 30 ~ "20-30",
    Age < 40 ~ "30-40",
    Age < 50 ~ "40-50",
    Age < 60 ~ "50-60",
    Age < 70 ~ "60-70",
    Age < 80 ~ "70-80",
    Age < 90 ~ "80-90",
    Age < 100 ~ "90-100"
  )) %>% 
  group_by(Age_Group) %>% 
  summarise(Total = n())

# Plotting the above table
political_extremism %>% 
  mutate(Age_Group = case_when(
    Age < 10 ~ "0-10",
    Age < 20 ~ "10-20",
    Age < 30 ~ "20-30",
    Age < 40 ~ "30-40",
    Age < 50 ~ "40-50",
    Age < 60 ~ "50-60",
    Age < 70 ~ "60-70",
    Age < 80 ~ "70-80",
    Age < 90 ~ "80-90",
    Age < 100 ~ "90-100"
  )) %>% 
  ggplot(aes(x = Age_Group, fill = Violent)) +
  geom_bar(position = "dodge") + 
  labs(title = "Distribution of Exposed Non-Violent and Violent Extremists in PIRUS by Age Group",
       x = "Age Group",
       y = "Number of Extremists",
       fill = "Violent\n(0 = No, 1 = Yes)") +
  scale_fill_manual(values = c("forestgreen", "firebrick")) +
  set_theme(new = theme_minimal())
```

```{r}
# plotting the number of exposed violent and non-violent extremists over time
political_extremism %>% 
  group_by(Violent, Year_Exposure) %>% 
  summarise(Total = n()) %>% 
  ggplot(aes(x = Year_Exposure, y = Total, color = Violent, group = Violent)) +
  geom_line() +
  labs(title = "Number of Exposed Violent and Non-Violent Extremists Over Time",
       x = "Year",
       y = "Number of Extremists",
       color = "Violent\n(0 = No, 1 = Yes)") +
  scale_color_manual(values = c("forestgreen", "firebrick")) + 
  set_theme(new = theme_minimal())
```

# Model Building

## Building an Unpenalized Logistic Regression Model

Splitting my data with an 80-20 split

```{r}
set.seed(2212003)
Violent <- political_extremism$Violent
political_extremism_exp <- political_extremism %>% 
  select(-Violent)

pirus_matrix <- model.matrix( ~ -1 + ., political_extremism_exp)
pirus2 <- as.data.frame(pirus_matrix)
pirus2 <- cbind(pirus2, Violent)

train_index <- sample(1:nrow(pirus2), size = 0.8 * nrow(pirus2))
train_data <- pirus2[train_index,]
test_data <- pirus2[-train_index,]
# Dropping factor levels not in the data to avoid multicollinearity
train_data <- droplevels(train_data)
test_data <- droplevels(test_data)

# Making train and test data
x_train <- model.matrix(Violent ~ ., train_data)[, -1]
y_train <- train_data$Violent

x_test <- model.matrix(Violent ~ ., test_data)[, -1]
y_test <- test_data$Violent
```

```{r}
# Up first: logistic regression with all variables
mod_max <- glm(Violent ~ ., train_data, family = "binomial")
summary(mod_max)
# There is perfect multicollinearity between Group_Membership Unknown, Group_Competition
```

I'll be looking at the **Area Under the Curve (AUC)** and **test accuracy** of my models to select my best model, and I will be **comparing the train accuracy to the test accuracy** of each model to determine if the models fit my data well.

```{r}
# Testing on the train data first
probs_max_train <- predict(mod_max, newdata = train_data, type = "response")
pred_max_train <- ifelse(probs_max_train > 0.5, 1, 0)
max_train_acc <- mean(pred_max_train == y_train)
max_train_acc

# Now on the test data
probs_max_test <- predict(mod_max, newdata = test_data, type = "response")

plot(roc(y_test, probs_max_test))
auc(roc(y_test, probs_max_test))

pred_max_test <- ifelse(probs_max_test > 0.5, 1, 0)
max_test_acc <- mean(pred_max_test == y_test)
max_test_acc
```

AUC = 0.8586 \~ 85.86%, Train Accuracy = 0.8949 \~ 89.49%, Test Accuracy = 0.7944 \~ 79.44%

## Building LASSO Regression Models with K-Fold Cross Validation

```{r}
# Training the model
cv_lasso <- cv.glmnet(x = x_train, y = y_train, nfolds = 10, alpha = 1, family = "binomial")
plot(cv_lasso)
lasso_min <- cv_lasso$lambda.min
lasso_1se <- cv_lasso$lambda.1se

lasso_min_mod <- glmnet(x_train, y_train, lambda = lasso_min, alpha = 1, family = "binomial")
lasso_1se_mod <- glmnet(x_train, y_train, lambda = lasso_1se, alpha = 1, family = "binomial")

coef(lasso_min_mod)

coef(lasso_1se_mod)
```

```{r}
# Testing my models, starting with min
probs_lmin_train <- predict(lasso_min_mod, newx = x_train, s = "lambda.min", type = "response")
pred_lmin_train <- ifelse(probs_lmin_train > 0.5, 1, 0)
lmin_train_acc <- mean(pred_lmin_train == y_train)
lmin_train_acc

probs_lmin_test <- predict(lasso_min_mod, newx = x_test, s = "lambda.min", type = "response")

plot(roc(y_test, probs_lmin_test))
auc(roc(y_test, probs_lmin_test))

pred_lmin_test <- ifelse(probs_lmin_test > 0.5, 1, 0)
lmin_test_acc <- mean(pred_lmin_test == y_test)
lmin_test_acc
```

The AUC is 0.9242 \~ 92.42%, the train accuracy is 0.8625 \~ 86.25%, and the test accuracy is 0.8481 \~ 84.81%.

```{r}
# Now for se1
probs_l1se_train <- predict(lasso_1se_mod, newx = x_train, s = "lambda.1se", type = "response")
pred_l1se_train <- ifelse(probs_l1se_train > 0.5, 1, 0)
l1se_train_acc <- mean(pred_l1se_train == y_train)
l1se_train_acc

probs_l1se_test <- predict(lasso_1se_mod, newx = x_test, s = "lambda.1se", type = "response")

plot(roc(y_test, probs_l1se_test))
auc(roc(y_test, probs_l1se_test))

pred_l1se_test <- ifelse(probs_l1se_test > 0.5, 1, 0)
l1se_test_acc <- mean(pred_l1se_test == y_test)
l1se_test_acc
```

The AUC of this model is 0.9305 \~ 93.05%, the train accuracy is 0.8421 \~ 84.21%, and the test accuracy is 0.8574 \~ 85.74%.

## Building Ridge Regression Models with K-Fold Cross Validation

```{r}
cv_ridge <- cv.glmnet(x = x_train, y = y_train, nfolds = 10, alpha = 0, family = "binomial")
plot(cv_ridge)

ridge_min <- cv_ridge$lambda.min
ridge_1se <- cv_ridge$lambda.1se

ridge_min_mod <- glmnet(x_train, y_train, lambda = ridge_min, alpha = 0, family = "binomial")
ridge_1se_mod <- glmnet(x_train, y_train, lambda = ridge_1se, alpha = 0, family = "binomial")

coef(ridge_min_mod)
coef(ridge_1se_mod)
```

```{r}
# Testing on train data
probs_rmin_train <- predict(ridge_min_mod, newx = x_train, s = "lambda.min", type = "response")
pred_rmin_train <- ifelse(probs_rmin_train > 0.5, 1, 0)
rmin_train_acc <- mean(pred_rmin_train == y_train)
rmin_train_acc

# Testing on test data with ROC/AUC
probs_rmin_test <- predict(ridge_min_mod, newx = x_test, s = "lambda.min", type = "response")

plot(roc(y_test, probs_rmin_test))
auc(roc(y_test, probs_rmin_test))

pred_rmin_test <- ifelse(probs_rmin_test > 0.5, 1, 0)
rmin_test_acc <- mean(pred_rmin_test == y_test)
rmin_test_acc
```

The AUC of this model is 0.9008 \~ 90.08%, the train accuracy is 0.8773 \~ 87.73%, and the test accuracy is 0.8166 \~ 81.66%

```{r}
# Train first
probs_r1se_train <- predict(ridge_1se_mod, newx = x_train, s = "lambda.1se", type = "response")
pred_r1se_train <- ifelse(probs_r1se_train > 0.5, 1, 0)
r1se_train_acc <- mean(pred_r1se_train == y_train)
r1se_train_acc

# Test second
probs_r1se_test <- predict(ridge_1se_mod, newx = x_test, s = "lambda.1se", type = "response")
# AUC/ROC
plot(roc(y_test, probs_r1se_test))
auc(roc(y_test, probs_r1se_test))

pred_r1se_test <- ifelse(probs_r1se_test > 0.5, 1, 0)
r1se_test_acc <- mean(pred_r1se_test == y_test)
r1se_test_acc
```

The AUC for this model is 0.9019 \~ 90.19%, the train accuracy is 0.8685 \~ 86.85%, and the test accuracy is 0.8222 \~ 82.22%.

This makes sense to me, as the ridge regularization process leaves in all of the parameters from my dataset, many of which were removed both in the standard logistic regression model and the LASSO logistic regression model.

Now, for the last models, I will build elastic nets.

## Building Elastic Net Regression Models with K-Fold Cross Validation

```{r}
cv_enet <- cv.glmnet(x = x_train, y = y_train, nfolds = 10, alpha = 0.5, family = "binomial")
plot(cv_enet)

enet_min <- cv_enet$lambda.min
enet_1se <- cv_enet$lambda.1se

enet_min_mod <- glmnet(x_train, y_train, lambda = enet_min, alpha = 0.5, family = "binomial")
enet_1se_mod <- glmnet(x_train, y_train, lambda = enet_1se, alpha = 0.5, family = "binomial")
```

Prediction time for Elastic Net

```{r}
# Train
probs_emin_train <- predict(enet_min_mod, newx = x_train, s = "lambda.min", type = "response")
pred_emin_train <- ifelse(probs_emin_train > 0.5, 1, 0)
emin_train_acc <- mean(pred_emin_train == y_train)
emin_train_acc

# Test
probs_emin_test <- predict(enet_min_mod, newx = x_test, s = "lambda.min", type = "response")

plot(roc(y_test, probs_emin_test))
auc(roc(y_test, probs_emin_test))

pred_emin_test <- ifelse(probs_emin_test > 0.5, 1, 0)
emin_test_acc <- mean(pred_emin_test == y_test)
emin_test_acc
```

The AUC of this model is 0.9239 \~ 92.39%, the train accuracy is 0.868 \~ 86.8%, and the test accuracy is 0.8407 \~ 84.07%

```{r}
probs_e1se_train <- predict(enet_1se_mod, newx = x_train, s = "lambda.1se", type = "response")
pred_e1se_train <- ifelse(probs_e1se_train > 0.5, 1, 0)
e1se_train_acc <- mean(pred_e1se_train == y_train)
e1se_train_acc

probs_e1se_test <- predict(enet_1se_mod, newx = x_test, s = "lambda.1se", type = "response")

plot(roc(y_test, probs_e1se_test))
auc(roc(y_test, probs_e1se_test))

pred_e1se_test <- ifelse(probs_e1se_test > 0.5, 1, 0)
e1se_test_acc <- mean(pred_e1se_test == y_test)
e1se_test_acc

# Use summary command
```

The AUC of this model is 0.927 \~ 92.7%, the train accuracy is 0.8504 \~ 85.04%, and the test accuracy is 0.8388 \~ 83.88%.

# Interpretation of Results

| Model | AUC | Train Accuracy | Test Accuracy | Accuracy Difference |
|----------------------------|-----------------------|-----------------------|--|--|
| Un-penalized Logistic Regression | 0.8586 (85.86%) | **0.8949 (89.49%)** | 0.7944 (79.44%) | 0.1005 (10.05%) |
| LASSO Logistic Regression ($\lambda = \hat{\lambda}$) | 0.9242 (92.42%) | 0.8625 (86.25%) | 0.8481 (84.81%) | 0.0144 (1.44%) |
| LASSO Logistic Regression ($\lambda = \bar{\lambda}$) | **0.9305 (93.05%)** | 0.8421 (84.21%) | **0.8574 (85.74%)** | +0.0153 (+1.53%) |
| Ridge Logistic Regression ($\lambda = \hat{\lambda}$) | 0.9008 (90.08%) | 0.8773 (87.73%) | 0.8166 (81.66%) |  |
| Ridge Logistic Regression ($\lambda = \bar{\lambda}$) | 0.9019 (90.19%) | 0.8685 (86.85%) | 0.8222 (82.22%) |  |
| Elastic Net Logistic Regression ($\lambda = \hat{\lambda}$) | 0.9239 (92.39%) | 0.8680 (86.8%) | 0.8407 (84.07%) |  |
| Elastic Net Logistic Regression ($\lambda = \bar{\lambda}$) | 0.927 (92.7%) | 0.8504 (85.04%) | 0.8388 (83.88%) |  |
